---
title: "Data Visualization with ggplot2"
author: "David Nganga"
date: today
format: 
 html: 
  self-contained: true
  embend-resources: true
  toc: true
execute:
  eval: true # Evaluate code
  echo: true # Show code
  error: true # Allows code execution with errors ignored
  warning: false # Suppress warnings
  message: false # suppress messages
editor_options:
  editor: visual
  chunk_output_type: inline
---

### Loading packages

```{r}
library(palmerpenguins)
library(tidyverse)
library(skimr)
library(knitr)
```

## **Structure of code**

The structure of the code for ggplot2 objects is of the form:

**Example: Palmer Penguins data**

➡️ First, let’s load the Palmer Penguins data set and explore some high-level characteristics of the data set using the code below:

```{r}
data(penguins, package = "palmerpenguins")


```

we can glimpse our data to start

```{r}
glimpse(penguins)
```

making a blank canvas for our visualization.

adding bill depth

```{r}
penguins |>
  ggplot(aes(x=bill_depth_mm)   )
```

adding bill length

```{r}
penguins |>
  ggplot(aes(x=bill_depth_mm), y=bill_length_mm  )
```

adding points to the plot:

```{r}
penguins |>
  ggplot(aes(x=bill_depth_mm,y=bill_length_mm))+
  geom_point()
```

➡️ Let’s color the points in our scatter plot based on the species of the penguin being Adelie, Chinstrap, or Gentoo, reproducing the plot below.

```{r}
penguins |>
  ggplot(aes(x=bill_depth_mm,
             y=bill_length_mm,
             color= species))+
  geom_point()
```

➡️ Next, add another layer to our ggplot to customize the title and subtitle using the labs() function to reproduce the plot below.

```{r}
penguins |>
  ggplot(aes(x=bill_depth_mm,
             y=bill_length_mm,
             color= species))+
  geom_point()+
  labs(title="Penguin bill length by bill depth",
       subtitle = "Dimensions for Adelie, chinstrap, and Gentoo Penguins")
```

➡️ Also customize the axis labels using the x and y options in the labs() function.

```{r}
penguins |>
  ggplot(aes(x=bill_depth_mm,
             y=bill_length_mm,
             color= species))+
  geom_point()+
  labs(title="Penguin bill length by bill depth",
       subtitle = "Dimensions for Adelie, chinstrap, and Gentoo Penguins",
       x="bill_depth(mm)",
       y="bill_length(mm)")

```

Adding caption

```{r}
penguins |>
  ggplot(aes(x=bill_depth_mm,
             y=bill_length_mm,
             color= species))+
  geom_point()+
  labs(title="Penguin bill length by bill depth",
       subtitle = "Dimensions for Adelie, chinstrap, and Gentoo Penguins",
       x="bill_depth(mm)",
       y="bill_length(mm)",
       caption = "Source:source")

```

➡️ Lastly, use a discrete color scale that is inclusive of viewers with common forms of color blindness by adding a scale_color_viridis_d() layer to the plot, creating the final plot below.

```{r}
penguins |>
  ggplot(aes(x=bill_depth_mm,
             y=bill_length_mm,
             color= species))+
  geom_point()+
  labs(title="Penguin bill length by bill depth",
       subtitle = "Dimensions for Adelie, chinstrap, and Gentoo Penguins",
       x="bill_depth(mm)",
       y="bill_length(mm)",
       color="species",
       caption = "Source:Station LTER / polmernguins packag") +
  scale_color_viridis_d() 

```

**Mapping vs. Setting**

```{r}
ggplot(penguins,
       aes(x = bill_depth_mm,
           y = bill_length_mm,
           size = body_mass_g)) + 
  geom_point()

```
