---
title: "Data Visualization with ggplot2"
author: "David Nganga"
date: today
format: 
 html: 
  self-contained: true
  embend-resources: true
  toc: true
execute:
  eval: true # Evaluate code
  echo: true # Show code
  error: true # Allows code execution with errors ignored
  warning: false # Suppress warnings
  message: false # suppress messages
editor_options:
  editor: visual
  chunk_output_type: inline
---

## Notebook Objectives

- Understanding the **code structure** for ggplot plots

- Understanding **layers of ggplot objects**

- Adding **labels and captions** of ggplot objects

- Understanding **aesthetics** of ggplot objects

- Knowing the difference between **mapping & setting aesthetics**

### Loading packages

```{r}
library(palmerpenguins)
library(tidyverse)
library(skimr)
library(knitr)
```

## The ggplot2 package

Its the most popular **Data Visualization** package in R.

## ggplot2 Structure of code

The structure of the code for ggplot2 objects is of the form:

" ggplot(data = \[dataset\],\
      mapping = aes(x = \[x-variable\],\
                    y = \[y-variable\] )) +\
  geom_xxx() +\
  other options "

## Example: Palmer Penguins data

The Palmer Penguins data set contains data on 344 penguins from 3 islands in [Palmer Archipelago, Antarctica](https://www.google.com/maps/place/Palmer+Archipelago/@-64.1463814,-62.1800963,7z/data=!3m1!4b1!4m5!3m4!1s0xbc78dd6dc38c572b:0xe609367aeed33087!8m2!3d-64.1957848!4d-62.0153384).

### Palmer penguins data summary

| Variable          | Description                     |
|-------------------|---------------------------------|
| species           | Species of the penguin          |
| island            | Island the penguin was found on |
| bill_length_mm    | Bill length (mm)                |
| bill_depth_mm     | Bill depth (mm)                 |
| flipper_length_mm | Flipper length (mm)             |
| body_mass_g       | Body mass (g)                   |
| sex               | Sex of the penguin              |
| year              | Year data was collected         |

➡️ First, let’s **load the Palmer Penguins data set** and explore some high-level characteristics of the data set using the code below:

```{r}
data(penguins, package = "palmerpenguins")


```

we can glimpse our data to start. (In R, **`glimpse()`** is a function from the **`dplyr`** package (part of the **`tidyverse`**) that gives you a quick, transposed preview of your dataset ).

```{r}
glimpse(penguins)
```

Activity1:

## ggplot2 data visualization steps

**visualizing the relationship between the bill depth of the penguins (mm) and the bill lengths (mm) of the penguins.**

[aesthetics](https://ggplot2.tidyverse.org/articles/ggplot2-specs.html) is used in data visualizations to differentiate between subgroups or to show additional dimensions of the data by mapping **characteristics to a variable** such as color, shape or size.

### Plot in layers

i\_ making a blank canvas for our visualization using the penguins data frame.

ii\_ adding bill depth to the **x-axis**.

```{r}
penguins |>
  ggplot(aes(x=bill_depth_mm)   )
```

iii\_ adding bill length to the **y-axis**.

```{r}
penguins |>
  ggplot( aes(x=bill_depth_mm, y=bill_length_mm  ) ) 
```

iv\_ adding **points** using **geom_point()** to the plot to create **scatter plot**. (Additional layers are added to a ggplot using the + operator).

```{r}
penguins |>
  ggplot( aes(x=bill_depth_mm, y=bill_length_mm) )+
  geom_point()
```

v\_ adding **color to the points** using **color()** in our scatter plot **based on the species of the penguin** being Adelie, Chinstrap, or Gentoo, reproducing the plot below.

```{r}
penguins |>
  ggplot(aes(x=bill_depth_mm,
             y=bill_length_mm,
             color= species))+
geom_point()
```

vi\_ adding another layer to our ggplot to customize the **Title and Subtitle** using the **labs()** function to reproduce the plot below.

```{r}
penguins |>
  ggplot(aes(x=bill_depth_mm,
             y=bill_length_mm,
             color= species))+
  geom_point()+
  labs(title="Penguin bill length by bill depth",
       subtitle = "Dimensions for Adelie, chinstrap, and Gentoo Penguins")
```

Activity2:

## ggplot2 customization

i\_ ➡️ Customizing the axis labels using the *x and y options* in the **labs() function**.

```{r}
penguins |>
  ggplot(aes(x=bill_depth_mm,
             y=bill_length_mm,
             color= species))+
  geom_point()+
  labs(title="Penguin bill length by bill depth",
       subtitle = "Dimensions for Adelie, chinstrap, and Gentoo Penguins",
       x="bill depth(mm)",
       y="bill length(mm)" )

```

ii\_ ➡️ Customize the legend title to be “Species” instead of “species”

```{r}
penguins |>
  ggplot(aes(x=bill_depth_mm,
             y=bill_length_mm,
             color= species))+
  geom_point()+
  labs(title="Penguin bill length by bill depth",
       subtitle = "Dimensions for Adelie, chinstrap, and Gentoo Penguins",
       x="bill depth(mm)",
       y="bill length(mm)",
       color= "Species")

```

iii\_ ➡️ Adding caption which may explain source of data

```{r}
penguins |>
  ggplot(aes(x=bill_depth_mm,
             y=bill_length_mm,
             color= species))+
  geom_point()+
  labs(title="Penguin bill length by bill depth",
       subtitle = "Dimensions for Adelie, chinstrap, and Gentoo Penguins",
       x="bill depth(mm)",
       y="bill length(mm)",
       color= "Species",
       caption = "Source:Station LTER / polmernguins packag" )

```

iii\_ ➡️ Lastly, use a **discrete color scale** that is inclusive of viewers with common forms of color blindness by adding a **scale_color_viridis_d()** layer to the plot, creating the final plot below.

```{r}
penguins |>
  ggplot(aes(x=bill_depth_mm,
             y=bill_length_mm,
             color= species))+
  geom_point()+
  labs(title="Penguin bill length by bill depth",
       subtitle = "Dimensions for Adelie, chinstrap, and Gentoo Penguins",
       x="bill_depth(mm)",
       y="bill_length(mm)",
       color="Species",
       caption = "Source:Station LTER / polmernguins packag") +
  scale_color_viridis_d() 

```

Congrats! 🎨 🎉 🥳 now created a visualization using ggplot2!

### Mapping vs Setting

**Mapping** in ggplot2 is the process of using aesthetics in a visualization to show **additional variables** from a data set

For example, in the plot below, **body_mass_g** is mapped to the **size** aesthetic.

i\_ mapping **body_mass_g** to the **size** aesthetic.

When we say **`body_mass_g`** is mapped to the size aesthetic, it means that the **visual size of each dot** on the plot is determined by **how heavy** that individual penguin is.

```{r}
ggplot(penguins,
       aes(x = bill_depth_mm,
           y = bill_length_mm,
           size = body_mass_g)) + 
  geom_point()

```

ii\_ mapping **flipper_length_mm** to the **alpha** aesthetic.

When we say **`flipper_length_mm`** is mapped to the alpha aesthetic, it means that **the transparency of each dot on the plot is determined by the length of that penguin's flippers**.

```{r}
ggplot(penguins,
       aes(x = bill_depth_mm,
           y = bill_length_mm,
           size = body_mass_g,
           alpha= flipper_length_mm)) + 
  geom_point()

```

### Setting

**Setting** in ggplot2 is the process of manually specifying aesthetics for a visualization based on fixed values for all observations.

i\_ set the value of aesthetic to 2

When we say **`size = 2`** sets the value of the aesthetic, it means you are manually forcing **every single dot on the plot to have a fixed, identical size**, completely ignoring any variables in the data.

```{r}
ggplot(penguins,
       aes(x = bill_depth_mm,
           y = bill_length_mm)) + 
  geom_point(size = 2)
```

ii\_ Reproduce the plot below by also setting the alpha aesthetic to be 0.50.

```{r}
ggplot(penguins,
       aes(x = bill_depth_mm,
           y = bill_length_mm)) + 
  geom_point(size = 0.50)
```

Question 1:

➡️ Is there any missing data? yes, the is a warning for missing values.

What is the plot doing with the missing values? Hint: consider using the skim() function from the skimr package to assess missingness.

```{r}
skim(penguins)
```

skim() function shows all the missing values in the data as shown above.

ggplot normally drops the missing values before plotting.

Question 2:

➡️ Reproduce the histogram of the penguin body masses displayed below using ggplot2. Hint: we can control the outline color of the bars in the histogram using the color aesthetic.

```{r}
penguins |>
  ggplot(aes(x = body_mass_g, 
             color = species)) +
  geom_histogram(fill = "white", position = "identity", alpha = 0.5, ) +
  labs(
    title = "Penguin body masses",
    x = "Body Mass (g)",
    y = "Frequency",
    color = "Species",
    caption = "Source: Palmer Station LTER / palmerpenguins package") +
  scale_color_viridis_d()
```

Question 3:

➡️ Reproduce the box plot of the penguin body masses displayed below using ggplot2. Hint: to suppress the y-axis text and ticks, add a theme(axis.text.y = element_blank(), axis.ticks.y = element_blank()) layer to the plot.

```{r}
penguins |>
  ggplot(aes(x = body_mass_g, 
             color = species, fill = species)) +
  geom_boxplot(alpha = 0.5) +
  labs(
    title = "Penguin body masses",
    x = "Body Mass (g)",
    y = NULL,
    color = "Species",
    fill = "Species",
    caption = "Source: Palmer Station LTER / palmerpenguins package"
  ) +
  scale_color_viridis_d() +
  scale_fill_viridis_d() +
  theme(
    axis.text.y = element_blank(), 
    axis.ticks.y = element_blank()
  )
```
